---
title: "README"
author: "Martin Skogholt"
date: "26 december 2018"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
This is a very fast implementation of the Naive Bayes classifier in R. It has the fastest execution time of any other Naive Bayes implementation in R. It's also the only implementation that makes it possible to use either a Bernoulli distribution or a multinomial distribution for the features. 

## Installation


## Development Version
To get the development version, you can install this package directly from Github
```{r eval=FALSE}
devtools::install_github("mskogholt/fastNaiveBayes")
```

## Usage
```{r eval=FALSE}
rm(list=ls())
library(fastNaiveBayes)

# Load Data
data("Titanic")

# Expand table function
expand.table <- function(x){
  tablevars <- expand.grid(rev(dimnames(x)))
  if (length(dim(x)) > 1) {
    ftablex <- ftable(x)
    counts <- as.vector(t(ftablex[, 1:ncol(ftablex)]))
  } else {
    counts <- as.vector(x)
  }
  expansion.index <- rep(1:nrow(tablevars), counts)
  newdat <- tablevars[expansion.index, , drop=FALSE]
  row.names(newdat) <- 1:nrow(newdat)
  revnames <- rev(names(newdat))
  newdat[, revnames, drop=FALSE]
}

# Expand table into individual level data frame
data_mat <- expand.table(Titanic)
data_mat <- data_mat[,c(4,1,2,3)]

# Create dummies from data frame
m <- ncol(data_mat)
col_counter <- m + 1
for(i in 2:m){
  vec <- data_mat[[i]]
  for(val in unique(vec)){
    data_mat[,col_counter]<-ifelse(vec==val,1,0)
    colnames(data_mat)[col_counter] <- paste0(colnames(data_mat)[i],"_",val)
    col_counter <- col_counter + 1
  }
}
# Only select survived and dummies
data_mat <- data_mat[,c(1,(m+1):ncol(data_mat))]

# Train naive bayes model
model <- fastNaiveBayes(data_mat[,2:ncol(data_mat)],data_mat[[1]], laplace = 1,
                        distribution = "bernoulli")
# Create predictions. Since predictions is made with data on which is trained this is the model fit
preds <- predict(model, data_mat[,2:ncol(data_mat)])

# Mean error of prediction
mean(data_mat[[1]]!=preds)
```

